# Constructing the smile

Payne decided to focus her analysis on three specific dates (or episodes):

- January 8, 2016 (Pre-Brexit)
- June 24, 2016 (Brexit)
- December 30, 2016 (Post-Brexit)

Roughly, the dates corresponded to six months ahead of the Brexit referendum, the week of the referendum, and the six months after the referendum.

In addition to the prices of the 25$\Delta$ options, she needed the prices of the 10$\Delta$ options. The equations below, analogous to the ones used to obtain the information from the 25$\Delta$ options, would prove handy for obtaining them: 

\[
\begin{aligned}
  \sigma_{10\Delta C} &= \sigma_{ATM} + BF_{10\Delta} + \frac{1}{2} RR_{10\Delta}\\
  \sigma_{10\Delta P} &= \sigma_{ATM} + BF_{10\Delta} - \frac{1}{2} RR_{10\Delta}\\
  \sigma_{90\Delta C} &= \sigma_{10\Delta P}
\end{aligned}
\]

Before performing the calculations Payne had to select the appropriate software. One option, which everybody at the IMF used, was Excel. One advantage was that free files were available in any dark corner in the IMF building. But there were horror stories about Excel: once one was hooked on it, it was impossible to abandon it. Some of her colleagues considered it even worse than being hooked on meth: users were saddled with the mental suffering and distress but did not experience any highs at all. 

`Matlab` was the other available software she could use but it was very expensive. The money was better spent buying a new Honda Accord fully equipped. She decided to use `R`, a free software environment for statistical computing and graphics, with many available state-of-the-art libraries [@R-base]. Furthermore, the savings could go into her defined-contribution retirement plan and allowed her to take the 50 option if she so desired.

## Selecting the data for specific dates

After begging the Internal Technocratic Diktat (ITD) Committee for permission to install the non-white-listed software, Payne finally got `R` and the `RStudio` IDE installed and running in her laptop. To read the data for the specific dates of the analysis she typed:

```{r echo=TRUE, out.width="100%", message=FALSE, warning=FALSE, paged.print=FALSE}

rm(list=ls())                                     # Clean up memory
filename = "2018_IET_Options_data.csv"            # Name of CSV data file
data = read.csv(filename, header=TRUE)            # Load datafile
data$Dates = mdy_hm(as.character(data$Dates))     # convert dates to Date class
rownames(data)=NULL                               # remove row names

# Specify dates for analysis

date01 = as.Date("2016-01-08 UTC")                 
date02 = as.Date("2016-06-24 UTC")
date03 = as.Date("2016-12-30 UTC")

# Create data frame this.data

this.data = rbind(
  data[which(data$Dates==date01),],
  data[which(data$Dates==date02),],
  data[which(data$Dates==date03),])

# Delete row names and change the names of the columns

rownames(this.data) = NULL
colnames(this.data) = c("Date","spot","forward","atm", "rr25","bf25",
                        "rr10","bf10","rf","rd","imp_rd")
```

## Calculation of implied volatilities (vols)

The following lines of code used the information from the risk reversals and the butterfly spreads to calculate the vols of the different calls and puts:

```{r echo=TRUE, message=FALSE, warning=FALSE, out.width="100%", paged.print=FALSE}
# Vols are in percent, expressed them as simple numbers
this.data$atm  = this.data$atm/100
this.data$rr25 = this.data$rr25/100
this.data$bf25 = this.data$bf25/100
this.data$rr10 = this.data$rr10/100
this.data$bf10 = this.data$bf10/100
this.data$rf   = this.data$rf/100
this.data$rd   = this.data$rd/100
this.data$imp_rd=this.data$imp_rd/100

# We will use this.data repeatedly
# Attach it to access its elements

attach(this.data)     

# Recover vols for different deltas and put them in the data frame

this.data$sigma10c = atm + bf10 + 0.5*rr10
this.data$sigma25c = atm + bf25 + 0.5*rr25
this.data$sigma75c = this.data$sigma25c - rr25
this.data$sigma90c = this.data$sigma10c - rr10
this.data$sigmaatm = atm

Tenor = 3/12   # Maturity of options, 3 months, in years
```

The tenor, or time to maturity of the option, was set to $\frac{3}{12} = 0.25$ years, to be consistent with the market convention for implied volatility, which is annualized.

## $\Delta_{ATM}$ market convention

So far, the only missing piece of information is the $\Delta_{ATM}$, or the $\Delta$ associated with the $ATM$ option. 
"How do I calculate that number?" Payne wondered aloud. As the reader can guess, the definition of the $ATM$ option depends on what convention the market uses. There are three different conventions:

- for retail products, the $ATM$ strike is the current spot rate. In this case:
\
\[
\begin{aligned}
K_{ATM} &= S\\
\Delta_{ATM} &= N \left(\frac{\log(F/S) + \frac{1}{2} \sigma_{ATM}^2 T}{\sigma_{ATM} \sqrt T} \right)
\end{aligned}
\]

- for emerging market currencies and/or options with maturities above one year, the $ATM$ strike is the forward rate, or the $ATMF$. In this case:
\[
\begin{aligned}
K_{ATM} &= F\\
\Delta_{ATM} &= \exp(-Rf \times T)N(\frac{1}{2}\sigma_{ATM} \sqrt T) 
\end{aligned}
\]

- for major currency pair and/or options with maturities of one year or less, the $ATM$ strike is the value of the exchange rate such that the call and the put have the same $\Delta$. In this case, $\Delta_{ATM} \simeq 0.5$ and the strike price $K_{ATM}$ is:
\
\[
\begin{aligned}
 K_{ATM} &= F \times \exp\left( 0.5 \sigma^2_{ATM}\times T \right)\\
 \Delta_{ATM} &= 0.5\times exp(-Rf \times T) \simeq 0.5 
\end{aligned}
\]

The pair GBP-USD required using the third market convention. Payne calculated the strike of the $ATM$ option and its $\Delta$:

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Calculate the strike of the ATM option
K_atm = forward*exp((0.5*(this.data$sigmaatm)^2)*Tenor)
deltaATM = 0.5*exp(-rf*Tenor)
```

## A rough first pass on the volatility smile

In the case of the GBPUSD, the $\Delta_{ATM}$ could be very well approximated by $0.5$. To check the data, Payne decided to plot of the volatility smile. First, she created the data frame with the information the plot required:

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Select only the vols for each delta
list_variables = c("sigma10c", "sigma25c", "sigmaatm", "sigma75c", "sigma90c")

# Read the data as a matrix
vol_data = t(as.matrix(subset(this.data, select=list_variables)))

# Group the deltas in a vector, to be used in the x-axis
delta_vector = c(0.10, 0.25, 0.5, 0.75, 0.9)

# Create the data frame for the chart
vol.smile  = data.frame(delta_vector, vol_data)   
rownames(vol.smile) = NULL
colnames(vol.smile) = c("Delta","PreBrexit","Brexit","PostBrexit")

```

and used `ggplot` to create the chart:

```{r echo=TRUE, out.width="100%", fig.asp=0.4, fig.cap = "Rough volatility smiles", eval=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(reshape2)
vol.data = melt(vol.smile, id="Delta")
ggplot(data=vol.data, aes(x=Delta, y=value, shape=variable)) +
   geom_point(aes(colour=variable), size=4) +
   labs(y="vol")
```


## A more refined approximation of the volatility smile

It is possible to extend the volatility smile over a wider range of $\Delta$ values using a polynomial approximation of second degree. The function `fit.Vol()` in the `auxFile.R` takes as input the observed vols and their corresponding $\Delta$s, and interpolates and extrapolates the vols for a wider range of $\Delta$ values:

```{r echo=TRUE, out.width="100%", fig.asp=0.4, eval=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Function fit.vol fits a polynomial of second degree to the volatility smile
fit.Vol = function(data.vol, data.delta,delta.range)
{
  poly.fit = lm(data.vol ~ poly(data.delta, 2, raw=TRUE))  

  # Use fitted polynomial to interpolate Delta-Vol Curve
  delta.square = delta.range*delta.range
  delta.interc = rep(1,length(delta.range))
  
  X = cbind(delta.interc, delta.range, delta.square)
  iVolInterpol = t(t(X)*poly.fit$coefficients)
  iVolInterpol = rowSums(iVolInterpol)

  return(iVolInterpol)
}
```

Payne used the `fitVol()` function to fit the volatility smile for the pre-Brexit, Brexit, and post-Brexit dates:

```{r echo=TRUE, out.width="100%", fig.asp=0.4, fig.cap = "Fitted volatility smiles", eval=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Get the data points from the vol.smile data frame
data.delta = vol.smile$Delta
data.vol  = vol.smile$PreBrexit

# Interpolation and extrapolation range
delta.range = seq(from=0.01, to = 0.99, by=0.005)

# Obtain the volatility smiles
smile.preBrexit = fit.Vol(vol.smile$PreBrexit, data.delta, delta.range)
smile.Brexit    = fit.Vol(vol.smile$Brexit, data.delta, delta.range)
smile.postBrexit = fit.Vol(vol.smile$PostBrexit, data.delta, delta.range)

# Group the extended volatility smiles in a data frame
smile.df = data.frame(delta.range, smile.preBrexit, smile.Brexit, smile.postBrexit)
rownames(smile.df) = NULL
colnames(smile.df) = c("Delta","preBrexit","Brexit", "postBrexit")

# Create the chart
library(reshape2)
smile.data = melt(smile.df, id="Delta")
ggplot(data = smile.data, aes(x=Delta, y=value, colour=variable)) + geom_line(size=1) +
  labs(y="vol")
```

Before going further in her analysis, Payne asked herself a question:

<div style = "width:auto; height:auto; padding-top:5px; 
padding-bottom:5px; padding-left:20px; padding-right:20px; font-size:90%; border:1px solid black">
Question: In the week of the Brexit referendum, does it make sense that higher $\Delta$s have higher implied volatility values?
</div>
